---
title: "Practical Fall 2019"
author: "Kasey Neff"
date: "10/28/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = "~/Documents/UMich_Bio201_F19/")
```

As with each lab, add "completed" to the Rmd filename; update the name in the author line above; update the location of the working directory as necessary. 

# Load packages
```{r Load packages, eval=TRUE, echo=TRUE, message=FALSE, warning=FALSE, results='hide'}
library(tidyverse)
library(readxl)
library(broom)
library(cowplot)
set.seed(7)
```

# Part 1 - concepts 
Type the answers to questions 1-5 in the space below. 

### Question 1

Given vector v1 below, will it be read as character or numeric? Why does this happen? [1 point]
```{r}
v1 <- c("one", "2", "three", "4")
```

<This would be read as a character because it is within quotation marks.> 

### Question 2

dataframe1 has 323 rows and dataframe2 has 345 rows. After an inner join the resulting data fame has 312 rows. Why does the result have fewer rows? [1 point]

<This result has fewer rows because you used an inner join. Inner joins delete any values/rows that contain unusable data such as NA values. The above data frames must've contained some NA values that were then deleted.> 

### Question 3

What are the three assumptions for a t-test? Explain what to do if the variance assumption is violated. Explain what to do if the normality assumption is violated. [3 points]

<The threee assumptions for a t-test are to check for normal distribution using the Shapiro test (want a medium to large p value), checking sample size is large enough and to check for equal vairances. If the variances are proven to not be equal (<0.10) than you must run a non parametric Willcox test. If the p value from the Shapiro test is small then we would want to create a histogram to visualize how normal the data distribution is. If it looks to be skewed or bimodal then we do not have a normal distribution.> 

### Question 4

You have a data frame (called butyrate_df) with the columns: "Participant_ID", "Study_week", "Semester", "Supplement_consumed", "Avg_butyrate". If this data frame is piped to each of the functions below, evaluate if the description of the results is true or false. [3 points]
```{r eval=FALSE, include=FALSE}
butyrate_df %>%
  filter(Semester == "Fall2019")
# Result1 = keeps only data from Fall 2019 semester

butyrate_df %>%
  filter(Avg_butyrate <= 47.0)
# Result2 = keeps only values equal to 47 mmol/kg 

butyrate_df %>%
  group_by(Semester) %>%
  summarize(New_column = median(Avg_butyrate))
# Result3 = new_column contains the median of each participant 
```

<Resutl1 = TRUE
Result2 = FALSE, keeps values less than or equal to 47.0
Result3 = FALSE, the new column includes the median Average butyrate values per Semester not per Participant_ID> 

### Question 5

something about NAs

The data frame (called breath_df) has the columns: "Participant_ID", "Study_week", "Semester", "Supplement_consumed", "Methane", "Hydrogen", "Carbon_dioxide". This data frame is used for the code below, explain what will happen to any NAs in each of the shown calculations, and how the NAs will effect the underlying data. [2 points]
```{r eval=FALSE, include=FALSE}
# Option 1
breath_df %>%
  filter(!is.na(Methane)) %>% 
  group_by(Participant_ID, Study_week) %>%
  summarize(Avg_methane = mean(Methane))

# Option 2
breath_df %>%
  group_by(Participant_ID, Study_week) %>%
  summarize(Avg_methane = mean(Methane, na.rm = TRUE))
```

<Option 1 = The NAs will not be removed from the mean Methane calculations. They will have an affect on the calculations. The exclamation mark present before the is.na function means to keep matchs to the function is.na for values of Methane. I believe this means to keep NA values.
Option 2 = The NAs will be removed from the mean Methane calculations. This will cause the resulting data frame to lose rows of data, but it also will not contain any missing values that may mess with/skew the mean. > 

# Part 2 - tidy data 

### Question 6

Find the data frame matching your unique name in the GitHub repository. Find the breath gas and taxa abundance data frames. Download and import all into RStudio and use these data frames to complete the rest of the practical. Name the dataframe to match your uniqname, name the breath gas data frame "breath", name the taxa abundance data frame "taxa". [1 point]
```{r}
# data import code #I changed it to lower case because that is how I prefer my data frames. It makes it easier for me.
kaseyn <- read_delim("Practial/raw_data/kaseyn.txt", 
                        delim = "\t", col_names = TRUE, trim_ws = TRUE, na = c("NA")) %>%
  rename_all(tolower)

breath <- read_delim("Practial/raw_data/breath.txt", 
                        delim = "\t", col_names = TRUE, trim_ws = TRUE, na = c("NA")) %>%
  rename_all(tolower)

taxa <- read_delim("Practial/raw_data/taxa.txt", 
                        delim = "\t", col_names = TRUE, trim_ws = TRUE, na = c("NA")) %>%
  rename_all(tolower)
```

Update column names in all files to match the course conventions. [1 point]
```{r}
#I did this in the section above... I hope that is okay.
```

View the supplement types in your semester (mine is Winter2015...). Are they all entered in the same style? If not update them to match. [1 point]
```{r}
glimpse(kaseyn)
kaseyn %>% 
  filter(supplement_consumed == "BRMPS")
```

### Question 7

What are the dimensions of each data frame? Write the results as a comment after the function. [1 point]
```{r}
dim(kaseyn) #155 rows, 9 columns 
dim(taxa) #46,881 rows, 6 columns
dim(breath) #2,668 rows, 5 columns
```

Select a join that will drop as few data points as possible. Join the data frame containing the participant data with the breath gas data. Name this "joined_df". [1 point]
```{r}
joined_df <- full_join(kaseyn, breath, by = c("participant_id", "study_week", "semester"), copy=FALSE) %>% 
  filter(study_week == "week1" | study_week == "week3")
```

What are the dimensions of the joined data frame? Write the results as a comment after the function. [1 point]
```{r}
dim(joined_df) #181 rows, 11 columns
```

### Question 8

Calculate the total SCFAs. [1 point]
```{r}
joined_df %>% 
  mutate(total_scfas = acetate_mmol_kg + butyrate_mmol_kg + propionate_mmol_kg)
```

Calculate the weekly mean of each SCFA of each participant. Calculate the weekly mean methane and hydrogen levels of each participant. [1 point]
```{r}
scfa_breath_means <- joined_df %>% 
  mutate(but_mean = mean(butyrate_mmol_kg, na.rm = TRUE),
         ace_mean = mean(acetate_mmol_kg, na.rm = TRUE),
         prop_mean = mean(propionate_mmol_kg, na.rm = TRUE),
         scfa_mean = (butyrate_mmol_kg + acetate_mmol_kg + propionate_mmol_kg/3)) %>% 
  mutate(methane_mean = mean(ch4, na.rm = TRUE),
         hydrogen_mean = mean(h2, na.rm = TRUE))
```

What are the dimensions of the data frame with the averages? Write the results as a comment after the function.
```{r}
dim(scfa_breath_means) #2603 rows, 17 columns
```

Join this data frame with the one called "taxa". Name the new data frame "all_data". What are the dimensions of this new data frame? [1 point]
```{r}
all_data <- full_join(scfa_breath_means, taxa) %>% 
  filter(study_week == "week1" | study_week == "week3") 

dim(all_data) #96718 rows, 21 columns
```

Save all data frames generated to your GitHub page. [1 point]
```{r}
write_delim(joined_df, path = "Practial/curated_data/joined_df.txt", delim = "\t")
write_delim(all_data, path = "Practial/curated_data/all_data.txt", delim = "\t")
write_delim(scfa_breath_means, path = "Practial/curated_data/scfa_breath_means.txt", delim = "\t")
write_delim(kaseyn, path = "Practial/curated_data/kaseyn.txt", delim = "\t")
write_delim(breath, path = "Practial/curated_data/breath.txt", delim = "\t")
write_delim(taxa, path = "Practial/curated_data/taxa.txt", delim = "\t")
write_delim(scfa_joined, path = "Practial/curated_data/scfa_joined.txt", delim = "\t")
```

# Part 3 - statistics & graphics 

Update all axes labels and figure titles so they are informative and easy to read. Upload final plots to GitHub. When specified, conduct the appropriate tests, and write the statistic (p, F, R value) as a comment. 

### Question 9

If you have more than one type of supplement consumed in your semester, filter for one type to answer Question 9. Plot each SCFA with the study week (weeks 1 and 3 only) on the x-axis. Make each study week a different color. Hint: you may have to use facets, or create individual plots and combine them into a multi-panelled figure. [3 points]
```{r}
But_plot <- all_data %>%
  filter(study_week == "week1" | study_week == "week3", 
         supplement_consumed == "BRMPS") %>% 
  ggplot(aes(x = study_week, 
             y = scfa_mean, 
             color = study_week)) + 
  geom_violin() + 
  xlab(NULL) + 
  ylab("SCFA (mmol/kg)") +
  labs(title = "The SCFA levels per study week with the use of BRMPS")
But_plot

save_plot(filename = "scfa_plot.pdf",
          plot = But_plot,
          nrow = 1, ncol = 2, 
          base_aspect_ratio = 1.1)

#I technically had two types for my semester (BRMPS and none), but I choose to filter for BRMPS because none is the control and also does not have a large enough sample size...
```

Based on the data presented in the plot you generated above, identify which statistical test you would need to determine if there was a change in SCFA concentration between weeks 1 and 3 of the study. Check the assumptions appropriate for the test. [3 points]
```{r}
# assumptions -- I would do a paired t-test
# large enough sample size? => Week 1 is 660 and Week 3 is 93. Both are large enough to continue.
all_data %>%
  group_by(supplement_consumed, study_week) %>% 
  filter(supplement_consumed == "BRMPS") %>% 
  summarise(sample_size = n()) 

# week1 normal distribution? => Skewed, with outliers everywhere, and the shapiro test has a p-value = < 2.2e-16, which means the distribution is very abnormal.
scfa_week1 <- all_data %>%
  filter(study_week == "week1",
         supplement_consumed == "BRMPS")
shapiro.test(scfa_week1$scfa_mean) #p-value = < 2.2e-16
scfa_week1_plot <- ggplot(scfa_week1, aes(x=scfa_mean)) +
  geom_histogram() +
  labs(title = "Histogram for Week 1 Distribution")
save_plot(filename = "scfa_week1_plot.pdf",
          plot = scfa_week1_plot,
          nrow = 1, ncol = 2, 
          base_aspect_ratio = 1.1)
# week3 normal distribution? => Skewed to the right, with a few outliers, and the shapiro test has a p-value = 3.636e-06.
scfa_week3 <- all_data %>% 
  filter(study_week == "week3",
         supplement_consumed == "BRMPS")
shapiro.test(scfa_week3$scfa_mean) #p-value = 3.636e-06
scfa_week3_plot <- ggplot(scfa_week3, aes(x=scfa_mean)) +
  geom_histogram() +
  labs(title = "Histogram for Week 3 Distribution")
save_plot(filename = "scfa_week3_plot.pdf",
          plot = scfa_week3_plot,
          nrow = 1, ncol = 2, 
          base_aspect_ratio = 1.1)

scfa_joined <- inner_join(x = scfa_week1, y = scfa_week3,
                    by = c("participant_id", "frequency", 
                           "semester", "supplement_consumed")) %>%
    rename(scfa_mean_wk1 = scfa_mean.x,
         scfa_mean_wk3 = scfa_mean.y) %>% 
    select(-starts_with("fract"), -starts_with("taxa"), -starts_with("category"),
           -starts_with("phylum"))

# variances?
var.test(x = scfa_joined$scfa_mean_wk1, 
         y = scfa_joined$scfa_mean_wk3, 
         alternative = "two.sided") #p-value = 0.0004861
# As a conclusion, the assumptions (alomst) all failed (non-normal distribution, unequal variances) therefore I should use the wilcox test.

wilcox.test(x = scfa_joined$scfa_mean_wk1,
            y = scfa_joined$scfa_mean_wk3, 
            alternative = "less", paired = TRUE, var.equal = FALSE)

# result? p-value = 1, the mean difference is exactly zero (they are the same).
```

Based on the results of the test, conclude if there was a difference in SCFAs between weeks. 

<I would conclude that there was no difference in SCFAs between week 1 and week 3 after taking BRMPS.> 

### Question 10

Is there a correlation between the abundace of hydrogen generators and breath hydrogen? Create a plot that shows the relationship and conduct the appropriate statistical test(s). [2 points]
```{r}
# plot
hydrogen_plot <- all_data %>%
  filter(category == "hydrogen_gen") %>%
  ggplot(aes(x = h2,
             y = fract)) + 
  geom_point() + 
  geom_smooth(method = "lm", 
              se = FALSE) + 
  xlab("Hydrogen breath levels") + 
  ylab("Relative abundance of H2 Generators") +
  labs(title = "Correlation between abundance of H2 generating bacteria and breath H2 levels")
save_plot(filename = "hydrogen_plot.pdf",
          plot = hydrogen_plot,
          nrow = 1, ncol = 2, 
          base_aspect_ratio = 1.1)

#sample size Hydrogen_gen = 1,941
all_data %>% 
  group_by(category) %>% 
  summarize(sample_size = n())

# normality: p-value = < 2.2e-16, skewed to the right
hydrogen_normality_check <- all_data %>%
  filter(category == "hydrogen_gen")
shapiro.test(hydrogen_normality_check$h2) 
hydrogen_histogram <- ggplot(hydrogen_normality_check, aes(x = h2)) + geom_histogram()

save_plot(filename = "hydrogen_histogram.pdf",
          plot = hydrogen_histogram,
          nrow = 1, ncol = 2, 
          base_aspect_ratio = 1.1)

#variances:
bartlett.test(h2 ~ fract, data = all_data)

#test: 
h2_correlation <- all_data %>%
  filter(category == "hydrogen_gen") %>% 
  lm(fract ~ h2, data = .) 
summary(h2_correlation)

# correlation result...
#p-value = 0.3953, not significant
# R-squared = -0.0001724, not significant
```

What can you conclude from these results?

<I can conclude that frm these results, the abundance of hydrogen generators does not correlate to hydrogen breath.> 

Is there a correlation between the abundace of hydrogen consumers and breath methane? Create a plot that shows the relationship and conduct the appropriate statistical test(s). [2 points]
```{r}
# plot
methane_plot <- all_data %>%
  filter(category == "hydrogen_consumers") %>%
  ggplot(aes(x = ch4,
             y = fract)) + 
  geom_point() + 
  geom_smooth(method = "lm", 
              se = FALSE) + 
  xlab("Methane breath levels") + 
  ylab("Relative abundance of hydrogen consumers") +
  labs(title = "Correlation between abundance of H2 consumng bacteria and breath CH4 levels")
save_plot(filename = "methane_plot.pdf",
          plot = methane_plot,
          nrow = 1, ncol = 2, 
          base_aspect_ratio = 1.1)
methane_plot

#sample size Hydrogen_consumers = 1,652
all_data %>% 
  group_by(category) %>% 
  summarize(sample_size = n())

# normality: p-value = < 2.2e-16, extremely skewed to the right
methane_normality_check <- all_data %>%
  filter(category == "hydrogen_consumers")
shapiro.test(methane_normality_check$ch4) 
methane_histogram <- ggplot(methane_normality_check, aes(x = ch4)) + geom_histogram()
methane_histogram
save_plot(filename = "methane_histogram.pdf",
          plot = methane_histogram,
          nrow = 1, ncol = 2, 
          base_aspect_ratio = 1.1)

#variances:
bartlett.test(ch4 ~ fract, data = all_data) 

#test: p-value =< 2.2e-16 and r-squared = 0.1642 
ch4_correlation <- all_data %>%
  filter(category == "hydrogen_consumers") %>% 
  lm(fract ~ ch4, data = .) 
summary(ch4_correlation)

# correlation result...
```

What can you conclude from these results?

<I can conclude that there is no correlation between hydrogen consumers and breath methane as the p-value is extremely small, and the R-squared value is small.> 

### Extra credit:

Is the abundace of hydrogen producers and consumers related? Create a plot, check assumptions, conduct the appropriate statistical test(s), and interpret the results. [1 point]
```{r}
# plot
hydrogen_extra_credit_plot <- all_data %>%
  filter(category == "hydrogen_gen" | category == "hydrogen_consumers") %>%
  ggplot(aes(x = fract, 
             y = hydrogen_gen, hydrogen_consumers)) + 
  geom_violin() 
hydrogen_extra_credit_plot
save_plot(filename = "hydrogen_extra_credit_plot.pdf",
          plot = hydrogen_extra_credit_plot,
          nrow = 1, ncol = 2, 
          base_aspect_ratio = 1.1)

#sample size Hydrogen_gen = 1,941 and hydrogen_consumers = 1,652
all_data %>% 
  group_by(category) %>% 
  summarize(sample_size = n())

# normality: p-value = < 2.2e-16, extremely skewed to the right
extra_credit_normality_check <- all_data %>%
  filter(category == "hydrogen_producers" | category == "hydrogen_consumers")
shapiro.test(extra_credit_normality_check$fract) 
methane_histogram <- ggplot(extra_credit_normality_check, aes(x = fract)) + geom_histogram()
methane_histogram

save_plot(filename = "methane_histogram.pdf",
          plot = methane_histogram,
          nrow = 1, ncol = 2, 
          base_aspect_ratio = 1.1)

# test(s) p-value =< 2.2e-16 and R-squared = 0.01644
hydrogen_extra_credit <- all_data %>%
  lm(fract ~ category, data = .) 
summary(hydrogen_extra_credit)
```

<They isn't strong correlation between the fraction abundance of hydrogen producers and consumers according to my analyses.> 


-----
end